---
title: "01_load"
author: 
  - "Asta Zeuner - s203544 - azeuner"
  - "Astrid Ginnerup - s203523 - AGinnerup"
  - "Emma Danø - s203564 - EDanoe"
  - "Andrea Kristensen - s193769 - AndreaMTK"
  - "Yayi Wang - s243554 - Yayi0117"
format: 
  html: 
    embed-resources: true
editor: visual
execute: 
  message: false
  warning: false
---

## Load libraries

```{r}
#| label: "load-libraries-01" 
#| echo: TRUE
#| eval: TRUE 
#| output: FALSE  

library("tidyverse")
library("table1")
library("here")
```

## Download data

Code for downloading the datasets we will be working with and saving in data/\_raw/ folder.

```{r}
#| label: "Download-protein-data-01"

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.protein.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

```{r}
#| label: "Download-survival-data-01"

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.survival.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

```{r}
#| label: "Download-clinical-data-01"

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.clinical.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

## Load data

read_tsv is used to load data files from data/\_raw and saving the files in objects.

```{r}
#| label: "load-data-01" 

protein_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.protein.tsv.gz"))
survival_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.survival.tsv.gz"))
clinical_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.clinical.tsv.gz"))

```

## Join meta data: Clinical and survival data

The clinical_and survival_data are joined using inner_join() to create a meta_data set. The joined meta_data is saved as a .tsv.gz file in data folder.

```{r}
#| label: "join-clinical-and-survival-data" 

meta_data <- inner_join(clinical_data, survival_data, by = c("sample" = "sample"))

data_dir <- here("data/")

write_tsv(x = meta_data, 
          file = str_c(data_dir, 
                       "01_meta_data_load.tsv.gz"))

```

## Joining meta_data and protein expression

The protein_data is pivoted longer using pivot_longer() and saved as a .tsv.gz file in data folder.

```{r}
#| label: "protein_data-pivot-longer"

peptide_long <- protein_data |> 
                pivot_longer(cols = -1, 
                             names_to = "sample", 
                             values_to = "expression")

data_dir <- here("data/")

write_tsv( x = peptide_long,
           file = str_c(data_dir, 
                  "02_peptide_expression_load.tsv.gz"))

```

meta_data and protein_long datasets are then joined using inner_join(). The resulting dataset is saved as a .tsv.gz file in data folder and will be the dataset used for cleaning the data.

```{r}
#| label: "Join-the-data-to-final"

# Join the datasets on the 'sample' column

expression_meta_data <- inner_join(meta_data, 
                         peptide_long, 
                         by = "sample")

data_dir <- here("data/")


write_tsv( x = expression_meta_data,
           file = str_c(data_dir, 
                  "03_expression_meta_data_load.tsv.gz"))

```
