---
title: "01_load"
author: "Group 21"
format: html
editor: visual
---

## Load libraries

```{r}
#| label: "load-libraries" 
#| echo: FALSE
#| eval: TRUE 
#| output: FALSE  

library("tidyverse")
library("table1")
library("here")
```

## Download data

```{r}

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.protein.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

```{r}

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.survival.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

```{r}

raw_dir <- here("data/_raw/")
data_file <- "TCGA-HNSC.clinical.tsv.gz"
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}

```

## Load data

```{r}
#| label: "load-data" 

protein_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.protein.tsv.gz"))
survival_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.survival.tsv.gz"))
clinical_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.clinical.tsv.gz"))

```

```{r}
patient_data <- full_join(clinical_data, survival_data, by = "sample")

patient_data |>
  sample_n(size = 10)
```

```{r}
write_csv(x = patient_data,
          file = "~/projects/group_21_project/data/patient_data.csv")
```

```{r}
protein_data <- t(protein_data)
```

```{r}
protein_data |>
  sample_n(size = 10)


## Combine data

```{r}
#| label: "join-clinical-and-survival-data" 

meta_data <- inner_join(clinical_data, survival_data, by = c("sample" = "sample"))

data_dir <- here("data/")

write_tsv(x = meta_data, 
          file = str_c(data_dir, 
                       "02_meta_data_load.tsv"))

```

```{r}
#| label: "pivot-longer-protein-expression" 

# Pivot the peptide expression data longer
peptide_long <- protein_data |> 
  pivot_longer(cols = -1, #exclude the first column of the dataset from being "pivoted"
               names_to = "sample", 
               values_to = "expression")

data_dir <- here("data/")

write_tsv(x = peptide_long, 
          file = str_c(data_dir, 
                       "03_protein_expression_data_load.tsv"))

## Combining the metadata

```{r}

meta_data <- full_join(clinical_data, survival_data, by = "sample")

data_dir <- here("data/")

write_tsv( x = meta_data,
           file = str_c(data_dir, 
                  "02_meta_data_load.tsv"))

```

## Subsetting the protein expression dataset

```{r}
#| label: "pivot-longer-protein-expression-dataset"

# Pivot the peptide expression data longer
peptide_long <- protein_data |> 
                pivot_longer(cols = -1, 
                             names_to = "sample", 
                             values_to = "expression")

data_dir <- here("data/")

write_tsv( x = peptide_long,
           file = str_c(data_dir, 
                  "03_peptide_expression_load.tsv"))


```

```{r}
#| label: "Join-the-data-to-final"

# Join the datasets on the 'sample' column


final_data <- inner_join(meta_data, 
                         peptide_long, 
                         by = "sample")

data_dir <- here("data/")


write_tsv( x = final_data,
           file = str_c(data_dir, 
                  "01_expression_meta_data_load.tsv"))


```
