---
title: "02_clean"
format: html
editor: visual
---

## Load libraries

```{r}
#| label: "load-libraries" 
#| echo: TRUE
#| eval: TRUE 
#| output: FALSE  

library("tidyverse")
library("table1")
library("here")
library("patchwork")
```

## Load dataset

```{r}
#| label: "load-data" 

expression_meta_data_clean <- read_tsv(file = here("data/03_expression_meta_data_load.tsv.gz"))


```

```{r}

test_dataframe <- meta_data |> 
  select(sample,
         cigarettes_per_day.exposures,
         alcohol_history.exposures,
         race.demographic,
         vital_status.demographic,
         vital_status.demographic,
         age_at_index.demographic,
         tissue_or_organ_of_origin.diagnoses,
         treatment_type.treatments.diagnoses)
  


```

-   add some of the columns from above to the cleaning below –\> then delete the above chunk

### Cleaning data

First we choose the wanted columns to work with in the wanted order and rename them:

-   remove the patient id (created under)

```{r}
#| label: "choosing-columns"

expression_meta_data_clean_1 <- expression_meta_data_clean |> 
  select(sample_id = sample,
         patient_id = submitter_id,
         gender = gender.demographic,
         vital_status = vital_status.demographic,
         overall_survival = OS,
         primary_site,
         pathologic_stage = ajcc_pathologic_stage.diagnoses,
         age_at_index = age_at_index.demographic,
         year_of_birth = year_of_birth.demographic,
         year_of_death = year_of_death.demographic,
         tissue_source_location = name.tissue_source_site,
         cigarettes_per_day = cigarettes_per_day.exposures,
         years_smoked = years_smoked.exposures,
         peptide_target,
         expression)
```

Then we trimmed the data-columns with sample_id and patient_id, to only have the necessary:

-   first remove the TCGA

-   create the patient_id and code_tissue_source_site by splitting the column

```{r}
#| label: "modify-sample-patient-id"

# Modify the sample_id and patient_id columns
expression_meta_data_clean_2 <- expression_meta_data_clean_1 |> 
  mutate(
    
    # Keep only the last part in sample_id and leave out TCGA-XX
    sample_id = sub("^TCGA-..-(.*)", "\\1", sample_id),
    
    # Keep only the ID part in patient_id and leave out TCGA-XX
    patient_id = sub("^TCGA-..-(.*)", "\\1", patient_id)
  )
```

## Write the clean dataset to a tsv.gz file

```{r}
#| label: "write-clean-data-to-tsv.gz"

data_dir <- here("data/")


write_tsv( x = expression_meta_data_clean_2,
           file = str_c(data_dir, 
                  "04_expression_meta_data_clean.tsv.gz"))


```

We want to check if values in the column "disease_type" are unique. If so, we confidently remove it

```{r}

clinical_data |> 
  distinct(race.demographic) 

```

Columns to keep:

Sample

cigarettes_per_day.exposures

alcohol_history.exposures

race.demographic

vital_status

age_at_index

tissue_or_organ

```{r}



```
