---
title: "06_analysis_2"
author:
- "Asta Zeuner - s203544 - azeuner"
- "Astrid Ginnerup - s203523 - AGinnerup"
- "Emma Danø - s203564 - EDanoe"
- "Andrea Kristensen - s193769 - AndreaMTK"
- "Yayi Wang - s243554 - Yayi0117"
  html: 
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

## Loading libraries

```{r}
#| label: "loading-libraries"

library("tidyverse")
library("broom")
```

## Loading data

```{r}
#| label: "load-data" 

expression_meta_data_aug <- read_tsv(file = here("data/04_expression_meta_data_clean.tsv.gz"))

```

## Selecting columns for linear modeling

Want to make a linear model of the peptide expression levels and the overall survival.

Also removing all missing values (NA) as they cannot be included in a linear model.

```{r}

expression_meta_data_analysis6 <- expression_meta_data_aug |> 
  select(overall_survival, 
         peptide_target,
         expression) |> 
   drop_na(expression)

expression_meta_data_analysis6
```

## Grouping the dataset by peptide_target and nesting.

```{r}
expression_meta_data_analysis6_nested <- expression_meta_data_analysis6 |> 
  group_by(peptide_target) |> 
  nest() 

expression_meta_data_analysis6_nested
```

Each peptide_target now has a dataframe containing the overall survival and the normalized gene expression for the patients. The dataset is grouped by the peptides

## Adding the linear model to the dataset

```{r}

expression_meta_data_analysis6_nested_model <- expression_meta_data_analysis6_nested |> 
  mutate(model_object = map(.x = data,
                            .f = ~lm(formula = expression ~ overall_survival,
                                   data = .x)))

expression_meta_data_analysis6_nested_model
```

```{r}

expression_meta_data_analysis6_nested_model |>
 filter(peptide_target == "1433BETA") |> # Replace "g2E09" with whatever was YOUR favourite gene!
 pull(model_object)
```

Computing the confidence intervals and p-value using tidy()

```{r}

expression_meta_data_analysis6_nested_model_tidy <- expression_meta_data_analysis6_nested_model |> 
  mutate(model_object_tidy = map(.x = model_object,
                                 .f = ~ tidy(conf.int = TRUE,
                                             conf.level = 0.95,
                                             x =.x)))

expression_meta_data_analysis6_nested_model_tidy
```

## Wrangling the data

```{r}

expression_meta_data_analysis6_estimates <- expression_meta_data_analysis6_nested_model_tidy |> 
  unnest(model_object_tidy)

expression_meta_data_analysis6_estimates

```
