---
title: "07_analysis_3"
author:
- "Asta Zeuner - s203544 - azeuner"
- "Astrid Ginnerup - s203523 - AGinnerup"
- "Emma Danø - s203564 - EDanoe"
- "Andrea Kristensen - s193769 - AndreaMTK"
- "Yayi Wang - s243554 - Yayi0117"
html: 
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

## **Loading libraries**

```{r}
#| label: "load-libraries" 
#| echo: TRUE
#| eval: TRUE 
#| output: FALSE  
#| 

library("tidyverse")
library("ggplot2")
library("here")

```

## Loading data

```{r}
#| label: "load-data" 

expression_meta_data_aug <- read_tsv(file = here("data/05_expression_meta_data_aug.tsv.gz"))

```

```{r}




barplot_vital_status_genders<-ggplot(data = expression_meta_data_aug,
       mapping = aes(x = gender,
                     fill = vital_status
                     ))+
  
  
  
  geom_bar(position = "fill"
            )+
  
  scale_fill_manual(
    values = c(
      "Alive" = "#eb4cbe",
      "Dead" = "#beeb4c"    
    )
  ) +
  

  
  
  labs(x = "Gender", 
       y = "Proportion",
       title = "Distribution of Gender and Vital Status",
       subtitle = "Proportional representation by Gender and Race",
       fill = "Vital Status"
       )+
  
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "right"
        )+
  theme_classic()


barplot_vital_status_genders  




ggsave(filename = "barplot_vital_status_genders.png",
       plot = barplot_vital_status_genders, 
       device = "png",
       path = "../results/images")







```

Facet wrap with ethnicity and survival

add absolute values

```{r}
barplot_vital_status_genders_facet_races<-ggplot(data = expression_meta_data_aug,
       mapping = aes(x = gender,
                     fill = vital_status
                     ))+
  geom_bar(position = "fill"
           )+
  
    scale_fill_manual(
    values = c(
      "Alive" = "#eb4cbe",
      "Dead" = "#beeb4c"    
    )
  ) +
  
  
  facet_wrap(~race)+
  labs(x = "Gender", 
       y = "Proportion",
       title = "Distribution of Gender and Vital Status Across Races",
          subtitle = "Proportional representation by gender and race",
       fill = "Vital Status") +
  
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text = element_text(size = 6, face = "bold"))+
  theme_classic()


barplot_vital_status_genders_facet_races

ggsave(filename = "barplot_vital_status_genders_facet_races.png",
       plot = barplot_vital_status_genders_facet_races, 
       device = "png",
       path = "../results/images")

```

```{r}

boxplot_years_smoked_vital_status_facet_pathological<-expression_meta_data_aug |> 
  drop_na(pathologic_stage) |> 
ggplot(mapping = aes(x = vital_status,
                     y = years_smoked))+
  
  
  
geom_boxplot(fill = "#eb4cbe",
             color = "#beeb4c")+
  
  

  
  facet_wrap(~pathologic_stage)+
  labs(x = "Vital Status", 
       y = "Years smoked",
       title = "Distribution of Years smoked and Vital Status Across pathological stages",
       fill = "Vital Status")+
  theme(plot.title = element_text(size = 10, hjust = 0.5))+
  theme_classic()


boxplot_years_smoked_vital_status_facet_pathological

ggsave(filename = "boxplot_years_smoked_vital_status_facet_pathological.png",
       plot = boxplot_years_smoked_vital_status_facet_pathological, 
       device = "png",
       path = "../results/images")

```

Apparently, no patients are recorded with years smoked = 0. We assume they are registered as NA although NA could also refer to the fact that the patients did not respond.

The 6 most differentially expressed genes between the two different vital status (dead/alive) sorted on different ages

```{r}

expression_of_6_significant_pep_and_vital_status<-expression_meta_data_aug |> 
  filter(peptide_target %in% c("Notch3", "FIBRONECTIN", "MERIT40_pS29","Enolase-2","PKC-a-b-II_pT638_T641","IMP3")) |> 
  
  ggplot(mapping = aes(x = age_at_index,
                       y = expression,
                       #fill = vital_status,
                       color = vital_status
                      ))+

  
  
  geom_point( shape = 8,
              size = 1)+
  
  scale_color_manual(
    values = c(
      "Alive" = "#eb4cbe",  
      "Dead" = "#beeb4c"  
    )
  ) +
  

  
  
  
  labs(x = "Age", 
       y = "Expression level",
       fill = "Vital Status",
       color = "Vital Status"
       )+
  
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.title = "Vital status",
        legend.position = "right"
        )+
  
  facet_wrap(~peptide_target)+
  
 
  
  theme_classic()


expression_of_6_significant_pep_and_vital_status

ggsave(filename = "expression_of_6_significant_pep_and_vital_status.png",
       plot = expression_of_6_significant_pep_and_vital_status, 
       device = "png",
       path = "../results/images")

```

With this plot we want to graphically/visually illustrate that some peptides expression are significantly associated with death. For example, upregulation of fibronectin is statistically significant for patiens who died. This can be a bit difficult to see from the plot. What is more evidently seen from the plot is the fact that MERIT40_pS29 is downregulated in dead people; the blue points cluster in the lower part of the plot.

Trying to think of another way to dislay the data in a less messy way:

```{r}

expression_of_6_significant_pep_and_vital_status<-expression_meta_data_aug |> 
  filter(peptide_target %in% c("Notch3", "FIBRONECTIN", "MERIT40_pS29","Enolase-2","PKC-a-b-II_pT638_T641","IMP3")) |> 
  
  ggplot(mapping = aes(x = age_interval_10_year,
                       y = expression,
                       fill = vital_status,
                       color = vital_status
                      ))+
  
  
  geom_point( shape = 8,
              size = 1)+
  
  
  
  labs(x = "Age", 
       y = "Expression level",
       fill = "Vital Status",
       color = "Vital Status"
       )+
  
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.title = "Vital status",
        legend.position = "right"
        )+
  
  facet_wrap(~peptide_target)+
  theme_classic()


expression_of_6_significant_pep_and_vital_status
```
