---
title: "08_analysis_4"
author:
- "Asta Zeuner - s203544 - azeuner"
- "Astrid Ginnerup - s203523 - AGinnerup"
- "Emma Danø - s203564 - EDanoe"
- "Andrea Kristensen - s193769 - AndreaMTK"
- "Yayi Wang - s243554 - Yayi0117"
html: 
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

## Loading libraries

```{r}
#| label: "loading-libraries"

library("tidyverse")
library("broom")

```

## Loading data

```{r}
#| label: "load-data" 

expression_meta_data_aug <- read_tsv(file = here("data/04_expression_meta_data_aug.tsv.gz"))

protein_data <- read_tsv(file = here("data/_raw/TCGA-HNSC.protein.tsv.gz"))

```

## Principle component coordinates

Transposing the dataframe so we can do a PCA of the peptide targets instead patients

```{r}
#| label: "transposing-protein-expression-dataframe" 

protein_data_trans <- protein_data |> 
  drop_na() |> 
  pivot_longer(col = -1,
               names_to = "sample",
               values_to = "expression") |> 
  pivot_wider(names_from = peptide_target,
              values_from = expression)

protein_data_trans
 
```

```{r}
#| label: "PCA" 
pca_analysis <- protein_data_trans |> 
  select_if(is.numeric) |> 
  drop_na() |> 
  prcomp(center = TRUE,
         scale = TRUE )
  
```

## Scree plot

```{r}
#| label: "scree-plot" 
#| fig-width: 7
#| fig-height: 4


Scree_plot <- pca_analysis |>
  tidy("eigenvalues") |> 
  mutate(percent = percent * 100) |> 
  ggplot(aes(x = PC,
             y = percent)) +
  geom_hline(yintercept = 0) +
  geom_col(colour = "orange",
           alpha = 0.5) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) +
  labs(title = "Scree Plot of PCA of Protein Expression",
       subtitle = "Protein Expression in Tissue from Head and Neck Cancer Patients",
       x = "Principle Component (PC)",
       y = "Percent")

ggsave(filename = "08_PCA_analysis_scree.png",
       plot = Scree_plot, 
       device = "png",
       path = "../results/images")

Scree_plot
```

From the scree plot it can be observed that a few of the peptides have a much higher percent variance compared to the rest of the peptides. Therefore further analysis are conducted.

## Adding the meta data to the PCA analysis

Joining the data frame used in the PCA with the meta_data, to be able to color based on vital status and pathological stage. We are using left join, as we only want the meta data for the patients that we have in our PCA.

```{r}
#| label: "joining-portein-expression-with-meta-data" 

protein_data_trans_metadata <- protein_data_trans |> 
  left_join(y = meta_data, 
             join_by(sample == sample)) 
```

Adding the metadata to the PC Analysis

```{r}
#| label: "augmenting-metadata-to-PCA-data" 

pca_analysis_aug <-  pca_analysis |> 
  augment(protein_data_trans_metadata)
```

Creating labels for the PCA plot, containing the variance percentage. The value is computed using tidy()

```{r}
#| label: "adding-variance-percentage-to-the-labels" 

pca_axis_labels <- pca_analysis |> 
  tidy("eigenvalues") |> 
  mutate(label = str_c("PC", PC, ", VE = ", round(percent*100,2), "%")) |> 
  pull("label")
```

## PCA plot

```{r}
#| label: "PCA-plot-vital-status" 
#| fig-width: 7
#| fig-height: 4


pca_plot_vital <- pca_analysis_aug |> 
  ggplot(aes(x = .fittedPC1, 
             y = .fittedPC2, 
             color = vital_status.demographic)) + 
  geom_point() +
  labs(x = pluck(pca_axis_labels,1),
       y = pluck(pca_axis_labels,2),
       title = "PC Analysis of Protein Expression" ,
       subtitle = "Protein Expression in Tissue from Head and Neck Cancer Patients",
       color = "Vital status") 

ggsave(filename = "08_PCA_analysis_PCA_vital.png",
       plot = pca_plot_vital, 
       device = "png",
       path = "../results/images")

pca_plot_vital
```

```{r}
#| label: "PCA-plot-pathological-stage" 
#| fig-width: 7
#| fig-height: 4

pca_plot_stage <- pca_analysis_aug |> 
  ggplot(aes(x = .fittedPC1, 
             y = .fittedPC2, 
             color = ajcc_pathologic_stage.diagnoses)) + 
  geom_point() +
  labs(x = pluck(pca_axis_labels,1),
       y = pluck(pca_axis_labels,2),
       title = "PC Analysis of Protein Expression" ,
       subtitle = "Protein Expression in Tissue from Head and Neck Cancer Patients",
       color = "Pathological stage") 

ggsave(filename = "08_PCA_analysis_PCA_stage.png",
       plot = pca_plot_stage, 
       device = "png",
       path = "../results/images")

pca_plot_stage
```

## Rotation matrix 

```{r}
#| label: "View-rotation-matrix-tibble"

pca_analysis |> 
  tidy(matrix = "rotation")
```

```{r}
#| label: "rotation-matrix"
#| fig-width: 7
#| fig-height: 7

arrow_style <- arrow(length = unit(0.03, "npc"),
                     ends = "first")

pca_analysis |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC", 
              names_prefix = "PC",
              values_from = "value") |> 
  ggplot(aes(x = PC1,
             y = PC2)) +
  geom_segment( xend = 0, 
                yend = 0, 
                arrow = arrow_style) +
  coord_fixed() + 
  theme_minimal() + 
  labs(x = pluck(pca_axis_labels,1),
       y = pluck(pca_axis_labels,2),
       title = "Rotation Matrix of PC Analysis of Protein Expression" ,
       subtitle = "Protein Expression in Tissue from Head and Neck Cancer Patients")
  
```
