---
title: "08_analysis_4"
author:
- "Asta Zeuner - s203544 - azeuner"
- "Astrid Ginnerup - s203523 - AGinnerup"
- "Emma Danø - s203564 - EDanoe"
- "Andrea Kristensen - s193769 - AndreaMTK"
- "Yayi Wang - s243554 - Yayi0117"
html: 
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

## Loading libraries

```{r}
#| label: "loading-libraries"

library("tidyverse")
library("broom")

```

## Loading data

```{r}
#| label: "load-data" 

expression_meta_data_aug <- read_tsv(file = here("data/04_expression_meta_data_aug.tsv.gz"))

```

## Principle component coordinates

Transposing the dataframe so we can do a PCA of the peptide targets instead patients

```{r}

protein_data_trans <- protein_data |> 
  drop_na() |> 
  pivot_longer(col = -1,
               names_to = "sample",
               values_to = "expression") |> 
  pivot_wider(names_from = peptide_target,
              values_from = expression)

protein_data_trans
 
```

```{r}
pca_analysis <- protein_data_trans |> 
  select_if(is.numeric) |> 
  drop_na() |> 
  prcomp(center = TRUE,
         scale = TRUE )
  
```

```{r}
pca_analysis |> 
  tidy("pcs")
```

```{r}

#make this plot prettier ###########################
pca_analysis |>
  tidy("pcs") |> 
  mutate(percent = percent * 100) |> 
  ggplot(aes(x = PC,
             y = percent)) +
  geom_hline(yintercept = 0) +
  geom_col(colour = "orange",
           alpha = 0.5) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) +
  labs(title = "Scree Plot of PCA of head and neck cancer 
       peptide expression")
```

Form the scree plot it can be observed that a few of the peptides have a much higher percent variance compared to the rest of the peptides. Therefore further analysis are conducted.

```{r}
protein_data_trans_metadata <- protein_data_trans |> 
  left_join(y = meta_data, 
             join_by(sample == sample)) 
```

```{r}

pca_analysis_aug <-  pca_analysis |> 
  augment(protein_data_trans_metadata)

```

ajcc_pathologic_stage.diagnoses

vital_status.demographic

```{r}
pca_analysis_aug |> 
  ggplot(aes(x = .fittedPC1, 
             y = .fittedPC2, 
             color = primary_site)) + 
  geom_point() 
```
